Haploid Selection, Sex Ratio Bias, and Transitions Between sex-determination systems

### 2017-09-14 LaTeX/Turnover.tex
some quick comments

\textcolor{red}{Change all  $\alpha^\Hermaphrodite$ to $(1+\alpha_{\Delta}^\Hermaphrodite)$.}
\textcolor{blue}{I've attempted this everywhere except in the recursions (S.1), which seem more natural with $\alpha$'s. I've run into trouble in equation S.8c,d -- I think we should check the Mathematica results to be sure we haven't made a typo. This also introduced an extra 1/2 in S.6c,d that might need to be explained.}

\textcolor{red}{Switch between $\chi$ and $\rho$ in all places because $\chi$ is used for double recombination events.}
\textcolor{blue}{I think we need a different variable given the $\rho$ terms in the characteristic polynomial. Also, where is $\chi$ used for double recombination? In other papers? Here $\chi$ is the probability of an odd number of cross-overs between the SDR and M loci.}

@@

\textcolor{red}{Change to $\hat{p}$ throughout as we assume that allele frequencies change slowly such that lambda is unaffected}
\textcolor{how will we differentiate between equilibrium $p$ and non-equilibrium $p$? or is the point that we can only say anything when at equilibrium?}

@@

%Sex ratio biases caused by gametic competition or meiotic drive have been shown to exert Fisherian sex-ratio selection on various autosomal \citep{Stalker:1961th,Smith:1975ft,Frank:1989vl,Hough:2013uo,Ubeda:2015fx, Otto:2015va} and sex-linked \citep{Kozielska:2010vm} modifiers.
Sex-ratio biases caused by haploid selection can facilitate heterogametic transitions between sex-determining systems. 
For instance, alleles favoured by haploid selection in males often become associated with the Y, which leads to a male-biased zygotic sex-ratio.
This male bias increases the potential for a neo-W to invade (Table \ref{tab:haplotype_growth}), which can equalize the sex-ratio \citep[e.g., see Figure \ref{fig:SexRatioBad}B, for related examples see][]{Kozielska:2010vm,Ubeda:2015fx}.
%We find that sex-ratio biases caused by haploid selection can also affect transitions between sex-determining systems (e.g., see $\zeta$ terms in Table \ref{tab:haplotype_growth}). 
%However, the neo-W may not equalize the sex-ratio (e.g., if the driving allele also drives in females or affects competition among eggs)
However, sex-ratio selection can be overwhelmed by additional selective effects (e.g., when a linked allele is beneficial for male diploids but detrimental for female diploids; Table \ref{tab:specialcases}), preventing the neo-W from invading. %and equalizing the sex ratio.
Indeed, transitions between sex-determining systems can even lead to stronger sex-ratio biases.
For example, where a neo-Y invades and is linked with a locus that experiences haploid selection in male gametes, the sex ratio evolves to become biased \citep[e.g., see Figure \ref{fig:SexRatioBad}A and step 1 in][]{Ubeda:2015fx}.
%For example, in an ancestral ZW system, haploid selection in males can allow a linked neo-Y to invade, despite the fact that it creates a male-biased sex ratio because the neo-Y becomes associated with the driven allele (Figure \ref{fig:SexRatioBad}A).
%\textcolor{blue}{and figure 1, where the end of one panel is the same as the beginning of another, however, this figure may change).}
%What we would like to stress is that sex-ratio selection alone cannot predict when new sex-determining systems can evolve.
Furthermore, with weak selection, we find that there is no difference in conditions allowing XY to ZW and ZW to XY transitions, indicating that sex chromosome transitions are not predominantly predicted by their effect on the sex-ratio (i.e., the sex-ratio bias created by male haploid selection facilitates the spread of a neo-W into an XY system the same way that male haploid selection drives the spread of a neo-Y into a ZW system with a 1:1 sex ratio). 
%An asymmetry can develop when sex-linkage is tight (e.g., Figure \ref{fig:Combination_Centimorgans} near -25cM and 25cM) but under most circumstances we do not predict asymmetry between XY to ZW and ZW to XY transitions despite the presence/absence of sex ratio selection. 
Thus, haploid selection can favour heterogametic transitions both via sex-ratio selection %\citep[as predicted by][]{Kozielska:2010vm} 
and via fitness effects of alleles that are associated with the neo-sex-determining allele, and these selection pressures are predicted to often be of roughly equal magnitude (unless linkage is tight). 
%\textcolor{red}{Maybe try to emphasise surprisingness in the paragraph more? e.g., how wrong you'd be if you assumed selection to equalise sex ratios, or it this implied?}
%\textcolor{blue}{nice, pretty subtle argument but I think that's a good job.}

@@

%\textcolor{blue}{similarly, this part doesn't discuss our new tight linkage results or our new perspective on sex ratios.}
%CONCLUSION
We have shown that tight sex-linkage and haploid selection can drive previously unexpected transitions between sex-determination systems.
In particular, both can select for neo-sex-determining loci that are more loosely linked. 
In addition, haploid selection alone can cause transitions analogous to those caused by purely sexually-antagonistic selection, eliminating the need for differences in selection between male and female diploids.
Perhaps counterintuitively, transitions involving haploid selection can be driven by sex-ratio selection or cause sex-ratio biases to evolve. 
We therefore argue that haploid selection should be considered as a pivotal factor in the evolution of sex-determination systems. 
Overall, our results suggest several new scenarios under which new sex-determination systems are favoured, which could help to explain why the evolution of sex-determination systems is so dynamic. 
%Further, we have shown the way in which transitions are affected by haploid selection is not intuitively obvious. 
%Firstly, sex-specific haploid selection affects turnovers between sex-determination systems in a manner that is qualitatively different from diploid sex-specific selection. 
%In particular, closer linkage between a sex-determining locus and a selected locus is not always favoured during heterogametic transitions when there is haploid selection. 
%Secondly, even though haploid selection is a source of zygotic sex-ratio biases, in our models Fisherian sex-ratio selection does not have good explanatory power in determining whether various sex-determination systems evolve. 
%This result is surprising given that sex ratios are ultimately determined via the sex-determination system, and leads us to the conclusion that three selective forces -- haploid, diploid, and sex-ratio selection -- should all be considered when exploring transitions between sex-determination systems.

@@

Secondly, haploid selection in females selects on neo-W haplotypes directly.
At equilibrium $(B)$, the fitness of female gametes under the ancestral sex-determining system is $w_{A}^\female$ such that the relative fitnesses of neo-W-$A$ and neo-W-$a$ haplotypes during female gametic competition are $w_{A}^\female/w_{A}^\female$ and $w_{a}^\female/w_{A}^\female$ (see terms in equation \ref{Binvasion}). 
Meiotic drive in females will also change the proportion of gametes that carry the $A$ versus $a$ alleles, which will be produced by heterozygous females in proportions $(1+\alpha_{\Delta}^\female)$ and $(1-\alpha_{\Delta}^\female)$, respectively. 
These terms are only associated with heterozygous females, i.e., they are found alongside $w_{Aa}^\female$.

Thirdly, haploid selection in males affects the diploid genotypes of females by altering the allele frequencies in the male gametes that female gametes pair with.
At equlibrium $(B)$, neo-W female gametes will mate with X-$A$ male gametes with probability $ w_A^\male (1+\alpha^\male_\Delta) / (2\bar{w}_{H}^\male)$ and Y-$a$ male gametes with probability $w_a^\male (1-\alpha^\male_\Delta) / (2\bar{w}_{H}^\male)$, where the $2\bar{w}_{H}^\male$ terms have been canceled in equation \eqref{Binvasion} (as mentioned above). 
Thus, for example, neo-W-$A$ haplotypes are found in $AA$ female diploids with probability $ w_A^\male (1+\alpha^\male_\Delta)/ (2\bar{w}_{H}^\male)$ (first term in square brackets in the numerate of equation \ref{Binvasion}a) and in $Aa$ female diploids with probability $w_a^\male (1-\alpha^\male_\Delta) / (2\bar{w}_{H}^\male)$ (see equation \ref{Binvasion}c and second term in square brackets in the numerator of equation \ref{Binvasion}a).

>>>>

\textcolor{red}{Change all  $\alpha^\Hermaphrodite$ to $(1+\alpha_{\Delta}^\Hermaphrodite)$.}
\textcolor{blue}{I've attempted this everywhere except in the recursions (S.1), which seem more natural with $\alpha$'s. I've run into trouble in equation S.8c,d -- I think we should check the Mathematica results to be sure we haven't made a typo. This also introduced an extra 1/2 in S.6c,d that might need to be explained.}
\textcolor{red}{hmm, not sure. This was an idea from Sally, I think in response to terms like $2 \alpha^\Hermaphrodite$ and $2(1-\alpha^\Hermaphrodite)$. It's possible that it makes other equations less easy to understand. My previous (not explained) logic was to use $w$ and $\alpha$ for strong selection and $s$, $t$, and $\alpha_\Delta$ for weak selection. Just to check, I should have written $\alpha^\Hermaphrodite=(1+\alpha_{\Delta}^\Hermaphrodite)/2$, maybe that's where the factors of $1/2$ come from?}

\textcolor{red}{Switch between $\chi$ and $\rho$ in all places because $\chi$ is used for double recombination events.}
\textcolor{blue}{I think we need a different variable given the $\rho$ terms in the characteristic polynomial. Also, where is $\chi$ used for double recombination? In other papers? Here $\chi$ is the probability of an odd number of cross-overs between the SDR and M loci.}
\textcolor{red}{Good point, not sure what's best for \textbf{the-term-formerly-known-as-$\rho$}. Yeah, Sally's written comment was something like ``$\chi$ is semi-standard notation for double recombination so it's use as a recombination rate might be confusing''.}

@@

\textcolor{red}{Change to $\hat{p}$ throughout as we assume that allele frequencies change slowly such that lambda is unaffected}
\textcolor{blue}{how will we differentiate between equilibrium $p$ and non-equilibrium $p$? or is the point that we can only say anything when at equilibrium?}
\textcolor{red}{I think Sally's point is that the eigenvalue analysis only makes sense if the $p$'s are changing slowly relative to the frequency of the modifier. Therefore, even when we don't substitute in any equilibrium frequencies, it might be safest to think of them as equilibrium allele frequencies. }

@@

%Sex ratio biases caused by gametic competition or meiotic drive have been shown to exert Fisherian sex-ratio selection on various autosomal \citep{Stalker:1961th,Smith:1975ft,Frank:1989vl,Hough:2013uo,Ubeda:2015fx, Otto:2015va} and sex-linked \citep{Kozielska:2010vm} modifiers.
Sex-ratio biases caused by haploid selection can facilitate heterogametic transitions between sex-determining systems. 
For instance, alleles favoured by haploid selection in males often become associated with the Y, which leads to a male-biased zygotic sex-ratio.
This male bias increases the potential for a neo-W to invade (Table \ref{tab:haplotype_growth}), which can equalize the sex-ratio \citep[e.g., see Figure \ref{fig:SexRatioBad}B, for related examples see][]{Kozielska:2010vm,Ubeda:2015fx}.
%We find that sex-ratio biases caused by haploid selection can also affect transitions between sex-determining systems (e.g., see $\zeta$ terms in Table \ref{tab:haplotype_growth}). 
%However, the neo-W may not equalize the sex-ratio (e.g., if the driving allele also drives in females or affects competition among eggs)
However, sex-ratio selection can be overwhelmed by additional selective effects (e.g., when a linked allele is beneficial for male diploids but detrimental for female diploids; Table \ref{tab:specialcases}), preventing the neo-W from invading. %and equalizing the sex ratio.
Indeed, transitions between sex-determining systems can even lead to stronger sex-ratio biases.
For example, where a neo-Y invades and is linked with a locus that experiences haploid selection in male gametes, the sex ratio evolves to become biased \citep[e.g., see Figure \ref{fig:SexRatioBad}A and step 1 in][]{Ubeda:2015fx}.
%For example, in an ancestral ZW system, haploid selection in males can allow a linked neo-Y to invade, despite the fact that it creates a male-biased sex ratio because the neo-Y becomes associated with the driven allele (Figure \ref{fig:SexRatioBad}A).
%\textcolor{blue}{and figure 1, where the end of one panel is the same as the beginning of another, however, this figure may change).}
%What we would like to stress is that sex-ratio selection alone cannot predict when new sex-determining systems can evolve.
Furthermore, with weak selection, we find that there is no difference in conditions allowing XY to ZW and ZW to XY transitions, indicating that sex chromosome transitions are not predominantly predicted by their effect on the sex-ratio (i.e., the sex-ratio bias created by male haploid selection facilitates the spread of a neo-W into an XY system the same way that male haploid selection drives the spread of a neo-Y into a ZW system with a 1:1 sex ratio). 
%An asymmetry can develop when sex-linkage is tight (e.g., Figure \ref{fig:Combination_Centimorgans} near -25cM and 25cM) but under most circumstances we do not predict asymmetry between XY to ZW and ZW to XY transitions despite the presence/absence of sex ratio selection. 
Thus, haploid selection can favour heterogametic transitions both via sex-ratio selection %\citep[as predicted by][]{Kozielska:2010vm} 
and via fitness effects of alleles that are associated with the neo-sex-determining allele, and these selection pressures are predicted to often be of equal magnitude when selection is weak. 
%\textcolor{red}{Maybe try to emphasise surprisingness in the paragraph more? e.g., how wrong you'd be if you assumed selection to equalise sex ratios, or it this implied?}
%\textcolor{blue}{nice, pretty subtle argument but I think that's a good job.}

@@

%\textcolor{blue}{similarly, this part doesn't discuss our new tight linkage results or our new perspective on sex ratios.}
%CONCLUSION
We have shown that tight sex-linkage and haploid selection can drive previously unexpected transitions between sex-determination systems.
In particular, both can select for neo-sex-determining loci that are more loosely linked. 
In addition, haploid selection alone can cause transitions analogous to those caused by purely sexually-antagonistic selection, eliminating the need for differences in selection between male and female diploids.
Perhaps counterintuitively, transitions involving haploid selection can be driven by sex-ratio selection or cause sex-ratio biases to evolve. 
We conclude that haploid selection should be considered as a pivotal factor driving transitions between sex-determination systems. 
Overall, our results suggest several new scenarios under which new sex-determination systems are favoured, which could help to explain why the evolution of sex-determination systems is so dynamic. 
%Further, we have shown the way in which transitions are affected by haploid selection is not intuitively obvious. 
%Firstly, sex-specific haploid selection affects turnovers between sex-determination systems in a manner that is qualitatively different from diploid sex-specific selection. 
%In particular, closer linkage between a sex-determining locus and a selected locus is not always favoured during heterogametic transitions when there is haploid selection. 
%Secondly, even though haploid selection is a source of zygotic sex-ratio biases, in our models Fisherian sex-ratio selection does not have good explanatory power in determining whether various sex-determination systems evolve. 
%This result is surprising given that sex ratios are ultimately determined via the sex-determination system, and leads us to the conclusion that three selective forces -- haploid, diploid, and sex-ratio selection -- should all be considered when exploring transitions between sex-determination systems.

@@

Secondly, haploid selection in females selects on neo-W haplotypes directly.
At equilibrium $(B)$, the fitness of female gametes under the ancestral sex-determining system is $w_{A}^\female$ such that the relative fitnesses of neo-W-$A$ and neo-W-$a$ haplotypes during female gametic competition are $w_{A}^\female/w_{A}^\female$ and $w_{a}^\female/w_{A}^\female$ (see terms in equation \ref{Binvasion}). 
Meiotic drive in females will also change the proportion of gametes that carry the $A$ versus $a$ alleles, which will be produced by heterozygous females in proportions $(1+\alpha_{\Delta}^\female)/2$ and $(1-\alpha_{\Delta}^\female)/2$, respectively. 
These terms are only associated with heterozygous females, i.e., they are found alongside $w_{Aa}^\female$.

Thirdly, haploid selection in males affects the diploid genotypes of females by altering the allele frequencies in the male gametes that female gametes pair with.
At equlibrium $(B)$, neo-W female gametes will mate with X-$A$ male gametes with probability $ w_A^\male (1+\alpha^\male_\Delta) / (2\bar{w}_{H}^\male)$ and Y-$a$ male gametes with probability $w_a^\male (1-\alpha^\male_\Delta) / (2\bar{w}_{H}^\male)$, where the $2\bar{w}_{H}^\male$ terms have been canceled in equation \eqref{Binvasion} (as mentioned above). 
Thus, for example, neo-W-$A$ haplotypes are found in $AA$ female diploids with probability $ w_A^\male (1+\alpha^\male_\Delta)/ (2\bar{w}_{H}^\male)$ (first term in square brackets in the numerator of equation \ref{Binvasion}a) and in $Aa$ female diploids with probability $w_a^\male (1-\alpha^\male_\Delta) / (2\bar{w}_{H}^\male)$ (see equation \ref{Binvasion}c and second term in square brackets in the numerator of equation \ref{Binvasion}a).
